---
title: "class05"
author: "Rita"
format: pdf
---

# Plotting in R

R has the multiple plotting and graphics systems. The most popular of which is **ggplot2**. 

```{r}

plot(cars)
```

ggplot is much more verbose than plot. Need to install the package.
```{r}
# install.packages("ggplot2")
library(ggplot2)

bb <- ggplot(cars, aes(speed, dist)) + geom_point()

bb

```

```{r}
bb +
  aes(x=speed, y=dist) +
  geom_point() +
  geom_smooth()
```


ggplot needs three things (data, aes(x,y)) + geom
```{r}
cc<-bb + geom_smooth(method='lm', se=FALSE)
```
```{r}
cc+ labs(title="Speed and Stopping Distances of Cars",
       x="Speed (MPH)", 
       y="Stopping Distance (ft)",
       subtitle = "Your informative subtitle text here",
       caption="Dataset: 'cars'") +
  geom_smooth(method="lm", se=FALSE) +
  theme_bw()
```


look at the top rows of the genes dataframe.

```{r}
url<-"https://bioboot.github.io/bimm143_W23/class-material/up_down_expression.txt"
genes<-read.delim(url)

head(genes)
```
plot the data

```{r}
ggplot(genes) + 
    aes(x=Condition1, y=Condition2,color=State) +
    geom_point() +
    labs(title="Gene distribution", subtitle="by states")
```

check the number rows and columns

```{r}
nrow(genes)
```
```{r}
ncol(genes)
```

```{r}
table(genes$State)
```

```{r}
round( table(genes$State)/nrow(genes) * 100, 2 )
```
This is the percentage of each categories

```{r}
p<-ggplot(genes) + 
    aes(x=Condition1, y=Condition2,color=State) +
    geom_point() +
    labs(title="Gene distribution", subtitle="by states")

p+scale_colour_manual( values=c("blue","gray","red") )
```

# Extra practice: The gap minder dataset
```{r}
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```
```{r}
library(dplyr)

gapminder_2007 <- gapminder %>% filter(year==2007)

```
```{r}
ggplot(gapminder_2007) +
  aes(x=gdpPercap, y=lifeExp) +
  geom_point(alpha=0.5)
```

```{r}
ggplot(gapminder_2007) + 
  geom_point(aes(x = gdpPercap, y = lifeExp,
                 size = pop), alpha=0.5) + 
  scale_size_area(max_size = 10)
```

```{r}
gapminder_1957 <- gapminder %>% filter(year==1957)

ggplot(gapminder_1957) + 
  aes(x = gdpPercap, y = lifeExp, color=continent,
                 size = pop) +
  geom_point(alpha=0.7) + 
  scale_size_area(max_size = 10) 
```

```{r}
gapminder_1957 <- gapminder %>% filter(year==1957 | year==2007)

ggplot(gapminder_1957) + 
  geom_point(aes(x = gdpPercap, y = lifeExp, color=continent,
                 size = pop), alpha=0.7) + 
  scale_size_area(max_size = 10) +
  facet_wrap(~year)
```

